% 0_acknows.tex
\chapter*{Acknowledgements}

The completion of this dissertation marks the end of a journey that would not have been possible without the support of a number of people, to whom I owe my sincere and heartfelt thanks.

To my mother, my deepest gratitude, for her unconditional support in every decision I have made, for her constant care, and for a mother's love that has never let me want for anything. She and my father have always set my education as one of their priorities. You have been, and will always be, my safe harbour and the strength that brought me this far.

To my father, for his steadfast support and presence throughout all my university years, and for never ceasing to believe in me.

To my grandmothers, strong and dignified women who gave shape and support to our whole family, and whose example of resilience I carry with me in everything I do.

To my brother, for encouraging me to pursue engineering and to always seek innovation, and for giving me the greatest joy in the world: my niece.

To all my friends, who accompanied me throughout my degree and my master's and filled these years with joy, good spirits, and moments I will treasure forever. Thank you for the unity with which we faced so many exams and projects. It was by sharing those battles that these years became unforgettable. A very special mention to Rita, one of my closest friends, who has guided my life and has always supported me during challenging periods. And to Rodrigo, for accompanying me and warmly supporting me throughout the last six months, in the new adventure that was undertaking this thesis in Liechtenstein.

To my supervisor at Hilti, Thibaut, for all his understanding, support, and motivation throughout the internship that gave rise to this thesis, and without whom this work would not have been possible.

To Professor Pedro Amorim, my supervisor at the Faculty of Engineering of the University of Porto, thank you for your invaluable guidance, knowledge, and support. I am honoured to consider you a professional and personal role model.

To all of you, my most heartfelt thank you.

I wish you a great reading!

\begin{flushright}
João Gonçalves
\end{flushright}
